---
title: "Lesson 9: Aggregate Demand and Aggregate Supply"
output:
  pdf_document: 
    toc: true
    toc_depth: 2
    number_sections: true
  html_notebook: default
---

```{r include = FALSE}
# refresh environment
rm(list = ls())

library(mdsr) # utility package for tidyverse
```

# Part One

# Part Two


## 3 Conclusions

1. No long run trade-offs between unemployment and inflation, because prices adjust
2. There is a short run trade-off
3. There are two types of Phillips Curves (SRPC, LRPC)

```{r}
# TODO insert SRPC vs LRPC graph here
```

## The Modern Phillips Curve

In 1973 and 1979 the United States economy encountered "oil shocks."
\[\pi = \pi^e - \omega(u-u_n) + \rho\]

### Phillips Curve with Adaptive Expectations

"What has the pattern been? I am going to predict according to that pattern."

\[\pi^e = \pi_{-1}\]
\[\pi = \pi_{-1}-\omega(u-u_n)+\rho\]

Two advantages:

1. Provides reason for sticky prices and sticky wages
    + $\pi$ may not fully adjust because people only look backwards
2. We can look at $\Delta\pi = \pi - \pi_{-1}$

\[\pi - \pi_{-1} = -q(u-u_n) + \rho\]

Assume $\rho = 0$

Therefor $\Delta\pi = 0$ if and only if $u = u_n$, so we call u~n~ NAIRU. This is
Non-Accelerating Inflation Rate of Unemployment.

# Part Three: Aggregate Supply Curves

Aggregate Supply Curves are the relationship between the output (Y) that firms
are willing to supply and inflation (price level $\pi$).

## Long Run Aggregate Supply Curve (LRAS)
```{r}
# TODO insert LR graph
```

"Input prices are going to change to what output prices are changing"

### Shifts

1. Increase in Productivity shifts LRAS right ($\Delta A$)
2. Increase in Capial, Labor, etc... shifts LRAS right ($\Delta inputs$)
3. Institutions (policy or such to create incentive for economic growth)
    + Property Rights, Rule of Law, education system, healthcare system...

## Short Run Aggregate Supply Curve (SRAS)

Take the Phillips curve and replace the unemployment gap ($u-u_n$) with the output
gap ($Y-Y^p$).

**Okun's Law:**

* For each percentage point Y is above Y^p^
* u is 1/2 percentage point below the natural rate u~n~
* $(u - u_n) = -0.5(Y = Y^p)$

SRAS equation (plug in Okun's law to PC):
\[\pi = \pi_{-1}+0.5\omega(Y - Y^p) + \rho\]
\[\pi = \pi_{-1}+ \gamma(Y - Y^p) + \rho\]

```{r}
# TODO make SRAS curve
```

### Shifts

1. Increase in expected inflation $\Delta\pi^e = \Delta\pi_{-1}$
2. Increase in price shock shifts curve up (left) ($\Delta\rho$)
3. The output gap ($\Delta(Y - Y^p)$)
    + If $Y\ne Y^p$ the we have $\Delta\pi$, so $\pi_{-1}$ changes












