---
title: "Lesson 4: Job Search"
output:
  pdf_document: 
    toc: true
    toc_depth: 2
    number_sections: true
  html_notebook: default
---

```{r include = FALSE}
# refresh environment
rm(list = ls())

library(mdsr) # utility package for tidyverse
```

## Key Terms

* U - Number of unemployed persons
* E - Number of employed persons
* V - Number of vacancies (job openings)
* u - Unemployment rate ($u = \frac{U}{LF}*100$)
* LF - Labor force ($LF = E + U$)
* $\theta$ - market tightness ($\theta = \frac{V}{U}$)
* Hires - Unemployed workers who become employed
* Separations - Employed workers who become unemployed
* Filled - Vacancies that become someone who is employed
* Separations - Employed workers who turn into a vacancy

# Matching Function

The flow of hires in an economy can be defined by a matching function $H(U, V)$
where an increase in unemployed or an increase in vacancies increases the number
of hires (positively correlated).

Example:
\[
H(U, V) = \gamma \sqrt{U} \sqrt{V}
\]
where  $\gamma$ is a coefficient that determines the matching efficiency.

## Job-Finding Rate ($f$)

The job-finding rate is defined as the probability that you find a job.

\[
f = \frac{H(U, V)}{U}
\]
Let us use our example:
\[
f = \frac{\gamma \sqrt{U} \sqrt{V}}{U}
\]
\[
f = \frac{\gamma U^{1/2} V^{1/2}}{U} = \gamma U^{-1/2} V^{1/2} = \gamma \frac{V^{1/2}}{U^{1/2}} = \gamma \sqrt{\theta}
\]


## Vacancy-Filling rate ($q$)

The vacancy-filling rate is defined as the probability that a vacancy is filled by
a firm within a given period.

\[
q = \frac{H(U, V)}{V}
\]
Let us use our example:
\[
q = \frac{\gamma U^{1/2} V^{1/2}}{V^{1/2}} = \gamma U^{1/2} V^{-1/2} = \gamma (\frac{U}{V})^{1/2} = \gamma \frac{1}{\sqrt{\theta}} = \gamma (\theta)^{-1/2}
\]

## A look at $f$ and $q$

Recall that $\theta$ is a measure of market tightness ($\frac{U}{V}$), so as $\theta$
increases the labor market becomes "tighter" (The ratio of vacancies to unemployed increases).
If the labor market is tight:

1. It is easy for unemployed workers to find a job.
2. It is difficult for a firm to fill a vacancy

The two ways this can happen is either an increase in vacancies or a decrease in unemployed
workers.

### Averages

* $\frac{1}{f}$ = Average duration of unemployment
* $\frac{1}{q}$ = Average time it takes to fill a vacancy
* $s = /frac{separations}{E}$ - separation rate
* $\frac{1}{s}$ = Average duration of employment

## Steady-State

Let's look back at the relationships between unemployment and employment:

1. Number of hires can be defined as $f * U$
    + $H(U, V) =  f * U$
2. Number of separations can be defined as $s * E$

Steady-state of unemployment would be considered the natural rate of unemployement:
\[
f * U = s * E
\]

We can solve for the natural rate of unemployment $u*$:
\[
u = \frac{U}{LF}
\]
\[
LF = E + U
\]
\[
f*U = E*s
\]
\[
f*U = (LF - U)*s
\]
\[
f*U = sLF - sU
\]
\[
sU + fU = sLF
\]
\[
\frac{(s + f)U}{LF} = s
\]
\[
u* = \frac{s}{(s + f)}
\]
Since we solved the job-finding rate for theta we can write:
\[
u* = \frac{s}{(s + f(\theta))}
\]

Which can be used to plot the a Bevridge curve:

```{r echo = FALSE}
gamma <- 0.5
s <- 0.05
uStar <- seq(from = 0, to = 10, by = 0.1)
theta <- ((1 / gamma) * ((s / uStar) - s)) ^ 2
data <- data.frame(theta = theta, uStar = uStar)

gamma2 <- 5
s3 <- 0.10
data$theta2 <- ((1 / gamma2) * ((s / uStar) - s)) ^ 2
data$theta3 <- ((1 / gamma) * ((s3 / uStar) - s3)) ^ 2

data %>%
  ggplot(aes(x = uStar, y = theta)) +
  geom_line() +
  geom_line(aes(y = theta2, color = "Increase gamama")) +
  geom_line(aes(y = theta3, color = "Increase s")) +
  xlab("Unempoyment Rate (u)") +
  ylab("Market Tightnes") +
  ggtitle("Bevridge Curve") +
  scale_x_continuous(limits = c(0, 1))
```


# Supply of Vacancies

A firm will open a vacancy if the marginal benefit of opening that vacancy is
greater than or equal to the marginal cost of opening that vacancy. The firm
earns a profit off a vacancy due to the output of the person they hire. The profit
that the firm earns is $\pi = y - w$ where y is the output of the worker (MPL) and w is 
their wage (in each period).

If we recall $\frac{1}{s}$ is the average duration of a job and q is the probability
of filling a vacancy (vacancy filling rate). This results in:
\[
MB = q * (y - w) * \frac{1}{s}
\]
Let us denote the cost of opening a vacancy (marginal cost) k. This means that any
given firm will continue to open vacancies until:
\[
k = q * (y - w) * \frac{1}{s}
\]

This is called the vacancy supply condition (free-entry condition).

If we go back to our example we can plug in our value for q:
\[
k = \gamma \frac{1}{\sqrt{\theta}} * (y - w) * \frac{1}{s}
\]

## Vacancy Supply curve


The vacancy supply curve is a function $\theta (w)$, which we can get by solving
our MB = MC equation of vacancies for $\theta$:

\[
k = \gamma \frac{1}{\sqrt{\theta}} * (y - w) * \frac{1}{s}
\]
\[
\sqrt{\theta} = \gamma \frac{1}{k} * (y - w) * \frac{1}{s}
\]
\[
theta(w) = (\frac{\gamma * (y - w)}{s * k})^2
\]

```{r echo = FALSE}
data$w <- seq(from = 0, to = 50, by = 0.5)
y <- 50
k <- 20
data$vs <- ((gamma * (y - data$w)) / (s * k))^2

data %>%
  ggplot(aes(x = w, y = vs)) +
  geom_line() +
  xlab("Wage (w)") +
  ylab("Market Tightnes") +
  ggtitle("Vacancy Supply Curve")
```

## Shifts in Vacancy Supply

Shifts can be seen within the vacancy supply function:

1. $\gamma$ +
2. y +
3. s -
4. k -

# Wage Formation (Wage-Setting Curve)

There could be a variety of factors that set wages:

* taxes
* y (output per worker)
* education
* minimum wage
* $\beta$ - bargaining power of unions
* b - unemployment benefits
* $\theta$ - market tightness

For us the Wage-Setting (WS) curve will be given. Example:
\[
w = \beta y + \beta k \theta + (1 - \beta) b ...
\]

```{r echo = FALSE}
data$ws <- 2 * data$w

data %>%
  ggplot(aes(x = w, y = ws)) +
  geom_line() +
  xlab("Wage (w)") +
  ylab("Market Tightnes") +
  ggtitle("Wage Setting Curve")
```


# Summary

We have three curves:

1. $w(\theta)$ wage setting VS
2. $\theta(w)$ vacancy supply WS
3. $u(\theta)$ Beveridge curve BC

```{r echo = FALSE}
# intecept guess
inter <- 33.5
data %>%
  ggplot(aes(x = w, y = ws)) +
  geom_line() +
  geom_line(aes(y = vs)) +
  geom_vline(xintercept = inter) +
  xlab("Wage (w)") +
  ylab("Market Tightnes") +
  ggtitle("Wage Setting Curve vs Vacancy Supply")
```


To find the unemployment rate associated with the equilbrium point, just find the
unemployment rate on the BC with the same market tightness. 


