---
title: "Lesson 5 Guide"
output:
  pdf_document: 
    toc: true
    toc_depth: 2
    number_sections: true
  html_notebook: default
---

```{r include = FALSE}
# refresh environment
rm(list = ls())

library(mdsr) # utility package for tidyverse
```

# Lesson 5: Consumption

## Consumption as part of GDP

Consumption accoutns for roughly 2/3 of GDP in the United States. This dictates
that it becomes a very important aspect around policy making. From 2001-2007
consumption grew by roughly 3% per year. During the great recession however
consumtion fell 0.25% in 2008 and 0.6% in 2009.

## Consumption Theories

In this lesson we look at consumer choice between consumption and saving (or 
future consumption).

* Theory 1: Intertemporal Choice (Irving Fisher)
    + we look at this
* Theory 2: Keynesian Consumption Function (Econ 104)
* Theory 3: Permanent Income Hypothesis (Friedman)
* Theory 4: Life-Cycle Hypothesis (Modigliani)

## Theory of Intertemporal Choice

### 3 steps to this theory

1. Define the Intertemporal Budget Constraint
2. Define the consumer preferences
3. Show the optimal consumption bundle

## Assumptions

1. Individuals live in a 2 period world
    + Period 1 vs Period 2
    + Current vs Future
    + Today vs Tomorrow
2. No Income/Wealth is left over
    + anything saved in period 1, is consumed in period 2
    + anything borrowed in period 1, is paid back in period 2
3. Savings rate = borrowing rate

## Terminology

Y~i~ = Income in period i (Y~1~ vs Y~2~)  
W~i~ = Wealth in period i  
C~i~ = Consumption in period i

(Y~i~ + W~i~) = i-Period's resources

With 2 time periods:

* (Y~1~ + W~1~) = current resources
* (Y~2~ + W~2~) = future resources
    + Combined they are lifetime resources
    
## Intertemporal Budget Line

```{r echo = FALSE}
y1 <- 100
w1 <- 50
y2 <- 50
w2 <- 200
r <- 0.05

c1 <- seq(from = 0, to = 500, by = 0.5)
c2 <- ((y1 + w1) * (1 + r) + (y2 + w2)) - (1 + r) * c1
IBL <- data.frame(c1 = c1, c2 = c2)

nlnbc1 <- y1 + w1
nlnbc2 <- y2 + w2

yintercept <- (1 + r) * (y1 + w1) + (y2 + w2)
xintercept <- (y1 + w1) + (y2 + w2) / (1 + r)

IBL %>%
  ggplot(aes(x = c1, y = c2)) +
  geom_smooth() +
  geom_point(aes(x = xintercept, y = 0)) +
  geom_point(aes(x = 0, y = yintercept)) +
  geom_point(aes(x = nlnbc1, y = nlnbc2)) +
  xlab("Current Consumption (C1)") +
  ylab("Future Consumption (C2)") +
  ggtitle("Intertemporal Budget Line") +
  scale_y_continuous(limits = c(0,500)) +
  geom_abline(slope = 1, intercept = 0) +
  annotate(geom = "text", x = xintercept - 50, y = 15,
           label = "Only Current Consumption") +
  annotate(geom = "text", x = 50, y = yintercept + 15,
           label = "Only Future Consumption") +
  annotate(geom = "text", x = nlnbc1 + 25, y = nlnbc2 + 25,
           label = "No Lending/No Borrowing")
```

### Points to observe

* Y-intercept = Only Future consumption (C~2~)
* X-intercept = Only Current consumption (C~1~)
* intserction of IBL and y = x is C~1~ = C~2~ (Consumption Smoothing)
* No-Lending/No-Borrowing Point
    + C~1~ = Y~1~ + W~1~
    + C~2~ = Y~2~ + W~2~
* Slope = -(1 + r)
    + r = real interest rate
    
## Math of IBL

Assumption 2: PVLC = PVLR

Present Value Lifetime Consumption (PVLC)  
\[
C_1 + \frac{C_2}{(1 + r)} + \frac{C_3}{(1 + r)^2} + \frac{C_4}{(1 + r)^3} + ... + \frac{C_n}{(1 + r)^{n-1}}
\]

Present Value Lifetime Resources (PVLR)
\[
(Y_1 + W_1) + \frac{(Y_2 + W_2)}{(1 + r)} + \frac{(Y_3 + W_3)}{(1 + r)^2} + \frac{(Y_4 + W_4)}{(1 + r)^3} + ... + \frac{(Y_n + W_n)}{(1 + r)^{n-1}}
\]

When n = 2 you get our IBL:
\[
C_1 + \frac{C_2}{(1 + r)} = (Y_1 + W_1) + \frac{(Y_2 + W_2)}{(1 + r)}
\]

Put it into `y = mx + b` form:
\[
C_2 = -(1 + r) * C_1 + ((Y_1 + W_1)(1 + r) + (Y_2 + W_2))
\]

## Shifts of the IBL
```{r echo = FALSE}
IBL$c2up <- IBL$c2 + 50
IBL$c2down <- IBL$c2 - 50

IBL %>%
  ggplot(aes(x = c1, y = c2)) +
  geom_smooth() +
  geom_smooth(aes(y = c2up, color = "More Resources")) +
  geom_smooth(aes(y = c2down, color= "Less Resources")) +
  xlab("Current Consumption (C1)") +
  ylab("Future Consumption (C2)") +
  ggtitle("Intertemporal Budget Line") +
  scale_y_continuous(limits = c(0,500))
```

Green (Shifts up/right):
* Increase in Y~1~, Y~2~, W~1~, W~2~

Red (Shifts down/left):
* Decrease in Y~1~, Y~2~, W~1~, W~2~

## Rotation of the IBL

```{r echo = FALSE}
r <- 0
IBL$c2slow <- ((y1 + w1) * (1 + r) + (y2 + w2)) - (1 + r) * IBL$c1
r <- 0.3
IBL$c2steep <- ((y1 + w1) * (1 + r) + (y2 + w2)) - (1 + r) * IBL$c1

IBL %>%
  ggplot(aes(x = c1, y = c2)) +
  geom_smooth() +
  geom_smooth(aes(y = c2slow, color = "Lower interest rates")) +
  geom_smooth(aes(y = c2steep, color= "Higher interest rates")) +
  xlab("Current Consumption (C1)") +
  ylab("Future Consumption (C2)") +
  ggtitle("Intertemporal Budget Line") +
  scale_y_continuous(limits = c(0,500))
```

An increase in r (real interest rate) correponds to a steeper slope and a decrease
in r corresponds to a shallower slope. This makes sense since the slope is -(1 + r)

## Preferences

The question: what combination of C~1~ and C~2~ would make an individual equally happy?

Use an indifference curve.

```{r echo = FALSE}
util <- 200
IBL$indiff <- (util / c1 ^ 0.5) ^ 2
IBL$indiffDouble <- (util * 2 / c1 ^ 0.5) ^ 2

IBL %>%
  ggplot(aes(x = c1, y = c2)) +
  geom_smooth() +
  geom_smooth(aes(c = c1, y = indiff, color = "IC")) +
  geom_smooth(aes(c = c1, y = indiffDouble, color = "IC * 2")) +
  geom_point(aes(x = xintercept, y = 0)) +
  geom_point(aes(x = 0, y = yintercept)) +
  geom_point(aes(x = nlnbc1, y = nlnbc2)) +
  xlab("Current Consumption (C1)") +
  ylab("Future Consumption (C2)") +
  ggtitle("Intertemporal Budget Line vs Indifference Curve (IC)") +
  scale_y_continuous(limits = c(0,500))
```

The indifference curve used in the example is at a utility of 200 for a cobb
douglas utility function of $U = C_1^{0.5} * C_2^{0.5}$

If "Marissa" always consumptions smooths, then you want to find the point on
the graph where C~1~ = C~2~

### To find C*:

1. Preference (C~1~ = C~2~)
2. IBL (PVLR = PVLC)

\[
C_1 + \frac{C_2}{(1 + r)} = (Y_1 + W_1) + \frac{(Y_2 + W_2)}{(1 + r)}
\]
Replace C~2~ with C~1~, since they are equal when smoothing
\[
C_1 + \frac{C_1}{(1 + r)} = (Y_1 + W_1) + \frac{(Y_2 + W_2)}{(1 + r)}
\]
\[
(1 + r) * C_1 + C_1 = (1 + r)(Y_1 + W_1) + (Y_2 + W_2)
\]
\[
C^* = \frac{(1 + r)(Y_1 + W_1) + (Y_2 + W_2)}{(2 + r)}
\]

What if "Sam" prefers to consume twice as much in the future period?

1. C~1~ = 0.5 * C~2~
2. IBL

\[
C_1^* = \frac{(1 + r)(Y_1 + W_1) + (Y_2 + W_2)}{(3 + r)}
\]

How does changes in r (real interest rate) affect consumption?

## Substitution Effect

If r increases, then C~1~ is more expensive for both borrowers and savers.

Therefore, C~1~ goes down and C~2~ goes up, for both types.

* Sub into "cheaper" good

## Income Effect
If r increases

* Borrower has less real income, so all consumption goes down
* Saver has more real income, so all consumption goes up

When looking at data, when the real interest rate increases, C~1~ decreases, so
the substitution effect is stronger than the income effect.



